Sex-Specific Evolution of the Meiotic Recombination Rate

### 2020-06-14 doc/MS_v3.Rmd
learning trying to convert to latex

---
title: 'Sex-Specific Evolution of the Meiotic Recombination Rate'
date: '`r Sys.Date()`'
author: "April L. Peterson, Bret Payseur"
output:
  html_document:
  word_document:
  pdf_document:
    includes:
      in_header: header.tex
always_allow_html: yes
bibliography: refs_Main_Report.bib
csl: elife.csl
---

@@

@@

Although meiotic recombination is required for successful gametogenesis in most species that reproduce sexually, the rate of crossing over varies among individuals. Differences in recombination rate between females and males are perhaps the most striking form of this variation. Existing data fail to directly address the extent to which recombination experiences similar evolutionary pressures. To fill this gap, we measured meiotic recombination in both sexes for a panel of house mouse cross three subspecies. Using inbred strains and single cell immunohistochemistry allowed us to place sex-specific observations within the same genetic background and meiotic context. Our results indicate highly discordant evolutionary patterns in the two sexes. Whereas male recombination rates show evidence of rapid evolution over short evolutionary timescales, female recombination rates measured in the same strains are mostly static. These results strongly indicate that house mouse has two genome wide recombination rates which display distinct evolutionary trajectories.

@@

@@

@@

@@

>>>>

---
title: 'Sex-Specific Evolution of the Meiotic Recombination Rate'
date: '`r Sys.Date()`'
author: "April L. Peterson, Bret Payseur"
output:
  pdf_document:
    includes:
      in_header: header.tex
  word_document:
  html_document:
always_allow_html: yes
bibliography: refs_Main_Report.bib
csl: elife.csl
---

@@

```{r global_options, include=FALSE}

knitr::opts_chunk$set(fig.width = 12, fig.height = 8, fig.path = 'figures/', echo = TRUE, warning = FALSE, message = FALSE)

@@

Although meiotic recombination is required for successful gametogenesis in most species that reproduce sexually, the rate of crossing over varies among individuals. **define genome-wide recombination rate**.

Differences in recombination rate between females and males are perhaps the most striking form of this variation. Existing data fail to directly address the extent to which recombination experiences similar evolutionary pressures. To fill this gap, we measured meiotic recombination in both sexes for a panel of house mouse cross three subspecies. Using inbred strains and single cell immunohistochemistry allowed us to place sex-specific observations within the same genetic background and meiotic context. Our results indicate highly discordant evolutionary patterns in the two sexes. Whereas male recombination rates show evidence of rapid evolution over short evolutionary timescales, female recombination rates measured in the same strains are mostly static. These results strongly indicate that house mouse has two genome wide recombination rates which display distinct evolutionary trajectories.

@@

Meiosis can be reduced to the expression of (2n -> 4n -> 2n -> 1n) which tracks the duplication of a diploid genome into haploid cell products. The meiotic program relies on crossovers and the process of recombination to ensure the correct separation of chromosomes. The total number of COs per cell (at the 4n stage) is equal to the genome wide recombination rate (gwRR). **The genome wide recombination rate is a cell based metric for genome wide recombination rate is tightly connected to organisms fitness through correct chromosome segregation and fertility.**

**at the fine scale / in regards to genetic variation**, In the indriect manner, the recombination rate regulates populations’ response to selection, and determine the fate of novel mutations by transferring beneficial mutations onto novel genetic backgrounds or by breaking linkage of negative mutations from beneficial genetic backgrounds (cite).  This process shapes the genomic patterns of genetic variation, with high recombination areas of genome having more nucleotide variation while areas of low recombination have decreased genetic variation (Begun Aquadro, Nachman, Payseur).

**but back to the genome-wide recombination rate**, The process of chromosome segregation is integral to the process of producing haploid cells and can be considered a constraint in regards to the evolution. When the genome wide recombination rate is framed within the single cell context it is closely connected to the  contraints which are imposed at the single cell level. Reflecting the importance of it's role in chromosome segreation is that the  number of haploid pairs is strong predictor for genome wide RR across large species(taxa) / divergence scales [@ottoPaysuer2019; @stapley_variation_2017]. (the number of crossovers per chromosome are highly conserved)<However, whie the rate of recombination is highly variable, the count of crossovers per chromosome is conserved between 1 to 3 (cite), suggesting that such thresholds/constraints may be behind the overall patterns.>

**connection of single cell based gwRR to thresholds imposed at the single cell level** (why is gwRR the better scale for evolutionary patterns?)

@@

Because we know the meiotic programs, we have expectations for where the thresholds(constraints) of the gwRR 
the rate will fall. First, the lower threshold is defined as an obligate crossover per homolog pair which promotes proper chromosome segregation [@nagaoka2012]. second (while there is less evidence,), the upper threshold maybe regulated/shaped/affected due to rates of ectopic recombination or the programmed DNA damage preceding crossovers act as constraints to the upper threshold of meiotic recombination (cite). Yet, Within these thresholds the genome wide recombination rate still varies across species and indiviuals. 
which is the subject of an open question in the field.  **the gwRR is the best metric for since both of these constraints are thought to act at the cell wide level.**

**Sex is one of the most notable axes along which individuals vary**. Sexual dimorphism in recombination rates is called heterochiasmy [@lenormand2003]. An understanding of how sex shapes the evolution of recombination cannot be achieved with available data. Comprehensive comparisons of female and male recombination rates usually come from outbred populations humans [@Kong2004;, @Kong2008; @Kong2014; @halldorsson2019], dog,  cattle [@ma2015_cattle; @Shen2018_cattle], (other sheep) sheep [@johnston2016_soay], and mouse (CC cite) in which the role of sex is confounded with the contributions of genetic variation. Under anisogamy where a species has distinct gametes, there is no first principle which would predict the evolution of sexually dimorphic recombination rates. Yet heterochiasmy, is commonly observed in dioecious species, suggesting that other meiotic traits which distinguish the gametes and their meiotic programs, for example symmetrical vs asymmetrical cell division, may impose selection for sexually dimorphic recombination rates. Although it is clear that the relationship between female and male recombination rates can differ among species, comparisons between and within closely related species are missing. Direct contrasts between females and males across a common, diverse set of genomic backgrounds would reveal whether the recombination rate evolves differently in the sexes.

<It’s important to note, – that the genome wide recombination rate can be decomposed – and connected to related traits (which also show conserved patterns of sexual dimorphism>

**understanding heterochiasmy and sexual dimorphism in meiosis have advanced since first being documented (morgan 1914) -- more than just the direction -- more evidence is accumulating that there are conserved patterns of sexual dimorphism in meiotic recombinatio - related to recombination rate**. Recently two specific features have been highlighted as conserved patterns of heterochiasmy:

@CahoonLibuda2019 show In species with meiotic chromatin (SC axis length) quantified for both sexes, longer chromosome axis generally has higher recombination rates. 
@sardell_sex_2020, A survey of 51 species shows there are conserved broad scale recombination landscape differences. Generally males have telomere-bias crossover placement and females have more uniform placement.

These observations raise the question of conservation in sexually patterns are maintained with the gwRR and it’s decomposed traits

**remember these points kinda distract from gwRR aspect**

@@

The House mouse complex arose from a recent radiation providing an opportunity to interrogate natural variation at short evolutionary scales, potentially narrowing the mutational space for identifying variants which would cause gwRR variation. Add divergence ranges. divergence spanning 0.5 mya to 5 mya.

Classical lab strains of mice have generated extensive knowledge central to meiosis and outcomes on recombination through extensive studies on the genes involved in the meiotic recombination pathway (cite) and previous crosses for understanding the genetic architecture of recombination rate variation @dumont2011house; @murdoch2010; @Wang2017island].

**Wild derived inbred strains generate the best comparison of females and males, besides the sex chromosomes, the mouse for each genome are almost identical**. Inbred strains enable one of the most direct comparison of both female and male --  versions of meiosis --enable use to conclude that sex is a primary factor --

Additionally due to their global distribution allows us to compare samples of natural diversity from a broad geographic range.  (wild derived inbred strains – will capture genetic diversity not represented in classical lab strains.)

While there are strains with Robertsonian translocations, in house mouse and related murid species it is possible to assemble sets of strains with identical karyotypes, 20 pairs of acrocentric chromosomes.

House mouse is well suited for single cell cytology approaches. The single cell level allows integration of data at a closer connection to the molecular pathway and meiotic program. In addition to the quantification of gwRR. This study quantifies precursors to crossovers, double strand breaks (DSBs), and meiotic chromosome morphology across stages of meiosis which would not be accessible to with genetic linkage data and crosses.

@@

  For the first time in many of these strains, We report a rare, direct, evolutionary comparison of recombination rate in females and males for this short evolutionary scale. We use rare strains from a broad range of geographic locations of the species territory.

The gwRR is made up of crossovers occurring on individual chromosomes within cells. Thus we quantified meiotic chromosome morphology (SC length) and placement of crossovers to comprise an approximate picture of the recombination landscape.

Our results indicate rapid male specific evolution of gwRR occurred in a subset of house mouse lines. We observed up to a 30% difference (translating into approximately ~7 more crossovers per cell), a surprising amount considering the short evolutionary time scale.
